---
title: "Vaccine Progress"
author: "Dan Gunn"
format: html
editor: visual
---

### About:

Summary of vaccination uptake.

The `echo: false` option disables the printing of code (only output is displayed).

### Packages

```{r}
library(tidyverse)
library(tidycensus)
library(skimr)
library(ggrepel)

```

```{r}
census_api_key(Sys.getenv("CENSUS_API_KEY"), install = TRUE)
```

### Import data 

```{r}
county_vax_raw <- read_csv("raw_data/COVID-19_Vaccinations_in_New_Jersey_by_County.csv", 
    col_types = cols(Date = col_date(format = "%m/%d/%Y")))

web_search_raw <- read_csv("raw_data/covid19.nj.gov Internal Search Data.csv")

ga_finder_raw <- read_csv("raw_data/GoogleAnalytics_covid19.nj.gov_finder.csv", 
    col_types = cols(Date = col_date(format = "%Y%m%d"), 
        `Avg. Session Duration` = col_time(format = "%H:%M:%S"), 
        `Avg. Time on Page` = col_time(format = "%H:%M:%S")))

ga_fullsite_raw <-  read_csv("raw_data/GoogleAnalytics_FullSitebyBrowserLanguage.csv", 
    col_types = cols(`Month of Year` = col_date(format = "%Y%m")), 
    skip = 1)

ga_fullsite_device_raw <- read_csv("raw_data/GoogleAnalytics_FullSitebyDeviceType.csv", 
    col_types = cols(`Month of Year` = col_date(format = "%Y%m")))
```

County list

```{r}
skim(county_vax_raw)
```

Contains an "unknown county" marker.

```{r}
county_list <- unique(county_vax_raw$Recip_County) |> sort()

length(county_list)

```

There are 136 instances of unknown county. I think this is a total count.

```{r}
county_vax_raw |> filter(Recip_County == "Unknown County") |> count()
#136 instances

county_vax_raw |> filter(Date == max(Date))|>
  filter(Recip_County != "Unknown County")|>
  summarize(sum(Series_Complete_Yes))
#zero doses
```

### Get county data for New Jersey

-what demos do I need? Make sure to use 2019 Census or ACS data.

-   Maybe get a 65+ age group population for each

-   Get a demographic breakdown for each county? Race / gender.. families?

```{r}
acs_v19 <- load_variables(2019, "acs5", cache = TRUE)
```

```{r}
#age
population_age_vars <- c(
  total_population = "B01003_001",
  median_age = "B01002_001",
  under_18_male = "B01001_003",    
  under_18_female = "B01001_027",
  age_65_plus_male = "B01001_020",
  age_65_plus_female = "B01001_044"
)

#race
race_ethnicity_vars <- c(
  white_non_hispanic = "B03002_003",
  black_or_african_american = "B03002_004",
  hispanic_or_latino = "B03002_012",
  asian = "B03002_006",
  american_indian_alaska_native = "B03002_005",
  two_or_more_races = "B03002_009"
)

#income
income_poverty_vars <- c(
  median_household_income = "B19013_001",
  per_capita_income = "B19301_001",
  individuals_below_poverty = "B17001_002",
  total_population_poverty = "B17001_001" 
)

#education
education_vars <- c(
  high_school_grad = "B15003_017",
  bachelors = "B15003_022",
  masters = "B15003_023",
  professional = "B15003_024",
  doctorate = "B15003_025"
)

# Employment
employment_vars <- c(
  employed_population_16_plus = "B23025_004",
  labor_force_total = "B23025_002",
  unemployed_population = "B23025_005"
)

#language
language_vars <- c(
  speak_only_english = "B16001_002",
  speak_spanish = "B16001_003"
)

#health insurance
health_insurance_vars <- c(
  uninsured_total = "B27001_005",
  total_population_health_insurance = "B27001_001" 
)

```

```{r}
all_vars <- c(
  population_age_vars,
  race_ethnicity_vars,
  income_poverty_vars,
  education_vars,
  employment_vars,
  language_vars,
  health_insurance_vars
)

nj_county_acs_long <- get_acs(geography = "county", 
              variables = all_vars, 
              state = "NJ", 
              year = 2019)
```

Pivot wider

```{r}
nj_county_acs_wide <- nj_county_acs_long |>
  #select(GEOID, NAME, variable, estimate, moe) |> 
  pivot_wider(
    names_from = variable, 
    values_from = c(estimate, moe)
  )

```

### Things to do.

Maybe a linear model to see where we're headed.

How to allocate unknown county data?

### Glimpse the county trends.

```{r}
county_vax_raw |>
  filter(Recip_County != "Unknown County") |>
  ggplot(aes(x = Date, y = Series_Complete_Pop_Pct, color = Recip_County)) +
  geom_line() +
  geom_label_repel(
    data = county_vax_raw |> 
      filter(Recip_County != "Unknown County") |>  
      group_by(Recip_County) |> 
      filter(Date == max(Date)),  
    aes(label = Recip_County),
    size = 4,
    nudge_x = 5,
    max.overlaps = 25
  ) +
  scale_x_date(
    limits = c(min(county_vax_raw$Date), max(county_vax_raw$Date) + 30),  
    date_labels = "%b %Y"  
  ) +
  theme_minimal()
```

```{r}
county_vax_raw |>
  filter(Recip_County != "Unknown County") |>
  ggplot(aes(x = Date, y = Series_Complete_Yes, color = Recip_County)) +
  geom_line() +
  geom_label_repel(
    data = county_vax_raw |> 
      filter(Recip_County != "Unknown County") |>  
      group_by(Recip_County) |> 
      filter(Date == max(Date)),  
    aes(label = Recip_County),
    size = 4,
    nudge_x = 5,
    max.overlaps = 25
  ) +
  scale_x_date(
    limits = c(min(county_vax_raw$Date), max(county_vax_raw$Date) + 30),  
    date_labels = "%b %Y"  
  ) +
  theme_minimal()
```

Cumberland county seems interesting.

Distributions

```{r}
county_vax_raw |>
   filter(Recip_County != "Unknown County") |>
  filter(Date > as.Date("2021-04-29"))|>
  ggplot(aes(y =reorder(Recip_County, Series_Complete_Pop_Pct), 
             x =Series_Complete_Pop_Pct))+
  geom_point()+
  labs(x = "Percent Complete",
       y = NULL)+
  theme_minimal()
```

### Pivoting longer so that I can see the county trends for different ages side by side.

```{r}
county_vax_long <- county_vax_raw |>
  filter(Recip_County == "Unknown County")

skim(county_vax_long)
```
